The sweet spot for local polygenic adaptation: a case study motivated by stickleback

### 2018-07-17 Stickleback_Paper.tex

Ninespine stickleback provide one of the most striking examples of local adaptation,
as the hemisphere-wide metapopulation includes both marine populations
and a large number of smaller freshwater populations
that have apparently repeatedly adapted to freshwater conditions
often using the same genetic basis.
In this paper, we use simulations motivated by stickleback populations
to examine what amounts of gene flow favor stable metapopulation polymorphism
with allele sharing,
and to further dissect the underlying dynamics.
We find that rapid, repeated adaptation 
using alleles maintained at low frequency by migration-selection balance
(the ``transporter hypothesis'')
occurs over a realistic range of intermediate rates of gene flow,
between slow, independent adaptation with low gene flow
and large migration load at high gene flow.
We do not see evidence for strong effects encouraging genomic clustering of causal alleles
or towards particular dominance coefficients.
$F_{ST}$ scans for adaptive alleles are more likely to succeed with higher rates of gene flow,
The results support existing theory of local adaptation,
and provide a more concrete look at a particular, empirically motivated example.

@@

\jgg{Starting a paper is awkward, help}
The rate and tempo of evolution remains largely unknown.
Today, we have found multiple examples as evidence of rapid evolution 
in tens of generations or fewer.
Biologists are now starting to address the complexity behind ecological speciation. 
\jgg{Peter and Bill: Could use help with examples of rapid evolution and the story leading up to stickleback right here}

Recently, multiple instances of similar (parallel) underlying genetic basis of rapid local adaptation has brought about 
questions surrounding the origins of adaptive alleles.
A empirical model is the alaskan populations of freshwater and marine Ninespine Stickleback fish.
In 1964, The Great Alaskan Earthquake caused an uplift of Middleton island and in turn,
introduced a group of freshwater ponds around the perimeter of the island. 
Quickly inhabited by the surrounding marine population of Stickleback,
\citet{lescak2015evolution} observed significant phenotypic changes in less than 50 years that appear to be 
parallel to freshwater stickleback that have been separated for over $13,000$ years.
In freshwater stickleback, the number of lateral plates are reduced
and the opercle shapes shows the same expansion of the dorsal region and reduction of the ventral region.
\jgg{Bill - more specifics about Bill, Suzie, Kristin  \& or Thom's papers .. ? }
These results leave us with questions surrounding the nature of rapid adaptation.
Does convergent evolution breed it's own solution (haplotype) for every new selective pressure, 
or can these solutions can be efficiently shared across multiple sub - populations facing similar selective pressures.

@@

% Overview
We varied the migration rate, $m$,
from $5 \times 10^{-5}$ to $5 \times 10^{-2}$ per individual,
i.e., between 0.01 and 10 migrants per year to and from each of the ten lakes.
This had a strong effect on many aspects of adaptation: 
including time to adaptation,
sharing of pre-existing freshwater adapted alleles,
and the population genetic signals left behind.
At the lowest migration rate, lakes adpated nearly independently,
while at the highest migration rate, the habitats are beginning to act like a single metapopulation
with substantial migration load.
We will now dissect what happens between these extremes.

Interestingly, the ability for separate populations to locally adapt to their own selective pressure was relatively unaffected 
until the highest rate of migration between the marine and freshwater environments.
All rates of historical introgression aside from the lowest, helped both the initial and introduced freshwater populations 
share pre-existing freshwater adapted alleles.
The sharing of alleles resulted in rapid adaptation ($\approx 100$ generations) of the introduced population (split from the marine population) to adapt to the freshwater environment.
We also found that larger amounts of migration allow for more statistical power and less false positives in the resulting population genetics data ($F_{st}$ per SNP across the genome).

@@

Local adaptation occurred in all simulations:
as shown in Figure \ref{fig:phenotype_ts2},
freshwater and marine populations diverged
until the trait means were close to the optimal values in each habitat. 
(Remaining simulations are shown in Figures \ref{phenotype_ts1} and \ref{phenotype_ts4}.)
Greater migration decreased adaptedness only weakly
except for at the highest migration rate (10 migrants per generation per lake),
where the mean trait value equilibrated at roughly half its optimal value 
(Figures \ref{fig:MeanPhenotype} and \ref{phenotype_ts4}).
With less than one migrant per generation per lake,
adaptation happened substantially faster
in the marine habitat than in the lakes.
The establishment of new alleles in the lakes is visible in Figure \ref{fig:phenotype_ts2}
as jumps in the mean phenotype curve;
in the two simulations these occur on a time scale of 100 (lower migration) to 1000 (higher migration)
generations,
and move the phenotype by an amount of order 1.
Trait variation within each population was small compared to the difference between populations,
with interquartile ranges of around XXX.\plr{TODO}
Across all parameter values, differences at around 16 commonly polymorphic sites 
(eight the shift the trait in each direction)
were responsible for most of the adaptive differences between freshwater and marine habitats.
\plr{ALSO TODO: CHECK THIS}

@@

As expected, increasing migration rate decreased differentiation between habitats.
As seen in Figure \ref{fig:Fst},
$F_{ST}$ between marine and freshwater habitats at neutral sites 
steadily declines as migration increases.
Local adaptation was still able to occur despite overall homogenization:
$F_{ST}$ between marine and freshwater habitats 
at sites with alleles affecting the trait (``effect mutations'')
was relatively constant across migration values.

@@

As expected,
lakes were able to adapt faster at higher migration rates
because greater mixing allowed them to share alleles
instead of developing their own genetic basis of adaptation.
As a first indication of this, Figure \ref{fig:Fst} shows that
$F_{ST}$ between the two sets of lakes at effect mutations decreased with migration rate.

To further investigate sharing of locally adaptive alleles, 
we investigate the ``pre-existing freshwater adapted alleles'',
defined for a particular generation
to be effect mutations above 50\% in at least one original lake and below 50\% in the marine habitat
in that generation.
Figure \ref{fig:MPFAI}A shows the distribution of the number of these alleles,
across generations,
as well as the average number of lakes each is found in.
As migration rates increase, the number of these alleles decreases steadily,
and each is concurrently found in a greater number of lakes.

>>>>

Ninespine stickleback provide one of the most striking examples of local adaptation,
as the hemisphere-wide metapopulation includes both marine populations
and a large number of smaller freshwater populations
that have apparently repeatedly adapted to freshwater conditions
often using the same genetic basis.
In this paper, we use simulations motivated by stickleback populations
to examine what amounts of gene flow favor stable metapopulation polymorphism
with allele sharing,
and to further dissect the underlying dynamics.
We find that rapid, repeated adaptation 
using alleles maintained at low frequency by migration-selection balance
(the ``transporter hypothesis'')
occurs over a realistic range of intermediate rates of gene flow,
between slow, independent adaptation with low gene flow
and large migration load at high gene flow.
This is mediated mainly by the total downstream influx of alleles \emph{(maybe??)}.
We do not see evidence for strong effects encouraging genomic clustering of causal alleles
or towards particular dominance coefficients.
$F_{ST}$ scans for adaptive alleles are more likely to succeed with higher rates of gene flow,
The results support existing theory of local adaptation,
and provide a more concrete look at a particular, empirically motivated example.

@@

\emph{intro outline:}
Start off with stickleback: repeated, rapid, parallel adaptation with obvious phenotypes.
But, most species don't seem to have such obviously structured polymorphism: 
what's special about stickleback?
The transporter hypothesis is a mechanism; what's required to make it work?
Other species adapt rapidly in similar ways:
e.g., Darwin's finches and island mice;
do these differ through not having such a big, connected reservoir of genetic variation?
or by not having such a large total population size of the ``other'' habitat?
Or, is this situation more common than we know, with less obvious phenotypes?
Also, can we actually tell what's happening from genetic data?

Recently, multiple instances of similar (parallel) underlying genetic basis of rapid local adaptation has brought about 
questions surrounding the origins and maintenance of genetic polymorphism
A empirical model is the Alaskan populations of freshwater and marine Ninespine Stickleback fish.
In 1964, The Great Alaskan Earthquake caused an uplift of Middleton island and in turn,
introduced a group of freshwater ponds around the perimeter of the island. 
Quickly inhabited by the surrounding marine population of Stickleback,
\citet{lescak2015evolution} observed significant phenotypic changes in less than 50 years that appear to be 
parallel to freshwater stickleback that have been separated for over $13,000$ years.
In freshwater stickleback, the number of lateral plates are reduced
and the opercle shapes shows the same expansion of the dorsal region and reduction of the ventral region.
\jgg{Bill - more specifics about Bill, Suzie, Kristin  \& or Thom's papers .. ? }
These results leave us with questions surrounding the nature of rapid adaptation.
Does convergent evolution breed it's own solution (haplotype) for every new selective pressure, 
or can these solutions can be efficiently shared across multiple sub - populations facing similar selective pressures.

@@

% Overview
We varied the migration rate, $m$,
from $5 \times 10^{-5}$ to $5 \times 10^{-2}$ per individual,
i.e., between 0.01 and 10 migrants per year to and from each of the ten lakes.
This had a strong effect on many aspects of adaptation,
including how fast adaptation occurred in each lake,
how much alleles were shared between lakes,
and the population genetic signals left behind.
At the lowest migration rate, lakes adpated nearly independently,
while at the highest migration rate, the habitats are beginning to act like a single metapopulation
with substantial migration load.
We will now dissect what happens between these extremes.

\emph{merge with the above}
Interestingly, the ability for separate populations to locally adapt to their own selective pressure was relatively unaffected 
until the highest rate of migration between the marine and freshwater environments.
All rates of historical introgression aside from the lowest, helped both the initial and introduced freshwater populations 
share pre-existing freshwater adapted alleles.
The sharing of alleles resulted in rapid adaptation ($\approx 100$ generations) of the introduced population (split from the marine population) to adapt to the freshwater environment.
We also found that larger amounts of migration allow for more statistical power and less false positives in the resulting population genetics data ($F_{st}$ per SNP across the genome).

@@

Local adaptation occurred in all simulations:
as shown in Figure \ref{fig:phenotype_ts2},
freshwater and marine populations diverged
until the trait means were close to the optimal values in each habitat. 
(Remaining simulations are shown in Figures \ref{fig:phenotype_ts1} and \ref{fig:phenotype_ts4}.)
Greater migration decreased adaptedness only weakly
except for at the highest migration rate (10 migrants per generation per lake),
where the mean trait value equilibrated at roughly half its optimal value 
(Figures \ref{fig:MeanPhenotype} and \ref{fig:phenotype_ts4}).
The establishment of new alleles in the lakes is visible in Figure \ref{fig:phenotype_ts2}
as jumps in the mean trait value;
in the two simulations these occur on a time scale of 100 (lower migration) to 1000 (higher migration)
generations,
and move the trait by an amount of order 1.
Trait variation within each population was small compared to the difference between populations,
with interquartile ranges of around XXX.\plr{TODO}
Across all parameter values, differences at around 16 commonly polymorphic sites 
(eight the shift the trait in each direction)
were responsible for most of the adaptive differences between freshwater and marine habitats.
\plr{ALSO TODO: CHECK THIS}

@@

As expected, increasing migration rate decreased differentiation between habitats.
As seen in Figure \ref{fig:Fst},
$F_{ST}$ between marine and freshwater habitats at neutral sites 
steadily declines as migration increases.
Local adaptation was still able to occur despite overall homogenization:
if computed using only sites with alleles affecting the trait (``effect mutations''),
$F_{ST}$ between habitats was relatively constant across migration values.

@@

The tenfold difference in speed of adaptation occurs because
at low migration rates, adaptation occurs independently in each lake,
and the marine habitat has ten times the influx of new alleles than any one lake.
In other words, 
greater mixing at higher migration rates allows lakes to share alleles
instead of developing their own genetic basis of adaptation.
As a first indication of this, Figure \ref{fig:Fst} shows that
$F_{ST}$ between the ``original'' and ``introduced'' sets of lakes 
at effect mutations decreased with migration rate.

To further investigate sharing of locally adaptive alleles
and the transporter hypothesis,
we investigate the ``pre-existing freshwater adapted alleles'',
defined for a particular generation
to be effect mutations above 50\% in at least one original lake and below 50\% in the marine habitat
in that generation.
Figure \ref{fig:MPFAI}A shows the distribution of the number of these alleles,
across generations,
as well as the average number of lakes each is found in.
As migration rates increase, the number of these alleles decreases steadily,
and each is concurrently found in a greater number of lakes.
